---
title: "StephanieChiang_607_2"
output: html_document
date: "2024-09-11"
params:
  pwd:
    label: "Please input password"
    value: ""
    input: password
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week 2 - R and SQL

### Parts 1 & 2: Build Table & Store data in SQL database

Connecting to the class Azure database took a little longer than expected at first. My machine is over 10 years old and still running MacOS 11 Big Sur (no longer supported for updates) so I had to find archived versions of MySQL and MySQL workbench.

Once connected and logged in, I manually input the table using the Workbench GUI, which was helpfully converted into SQL code too.

### Part 3: Transfer data from SQL database to R dataframe

To begin with, I was protecting the db password with R Studio's built-in password input function, but it broke when knitting the document and execution of the chunk halted with errors:

> pw <- .rs.askForPassword("Database Password:")

Confirmed [here](https://github.com/quarto-dev/quarto-cli/issues/2269) that this is a known R Studio issue that they have no intention of ever fixing. 

Instead, I used a 'parameterized report' by updating the top of the r markdown file, and then selecting 'Knit with Parameters':

```
title: "StephanieChiang_607_2"
output: html_document
date: "2024-09-11"
params:
  pwd:
    label: "Please input password"
    value: ""
    input: password
```

```{r sql-connection, warning=FALSE}
library("RMySQL")

sc_connection <- dbConnect(MySQL(),
                        user="stephanie.chiang02",
                        password=params$pwd,
                        dbname="stephanie.chiang02",
                        host="cunydata607sql.mysql.database.azure.com",
                        port=3306)

dbListTables(sc_connection)

movies_2024 <- dbReadTable(sc_connection, "movies_2024")
```

### Part 4: Missing data strategy 

First, I converted the column to numeric. From there, any strategy chosen to handle missing values would depend on the aim.

``` {r missing-data-examples}
movies_2024$score <- as.numeric(as.character(movies_2024$score))

# For example, if I want an overall consensus of the quality of movies this year:
# strip the values using na.rm=TRUE when calculating the median
total_median <- median(movies_2024$score, na.rm=TRUE)
print(total_median)

# Or another example, counting the viewership/popularity of a movie:
# use is.na() to remove from the count
challengers_viewership <- nrow(
  movies_2024[movies_2024$title == "Challengers" & !is.na(movies_2024$score),]
)
print(challengers_viewership)

# In the future, for a recommender, a subset of NA values could be created
# perhaps even for each reviewer
mom_unwatched <- movies_2024[movies_2024$reviewer=="Mom" & is.na(movies_2024$score),]
library(knitr)
kable(mom_unwatched)
```

### Conclusion

```{r disconnect}
# Close the connection once finished, to conserve system resources
# (helpful when your laptop is as old as mine)
dbDisconnect(sc_connection)
```
