---
title: "Week 11 Extra Credit Assignment"
author: 'Stephanie Chiang'
output: html_document
date: "2024-11-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Global Baseline Estimate 

The Global Baseline Estimate (GBE) is a non-personalized prediction algorithm that uses a global average across all users and items as a baseline estimate. Bias is then mitigated by calculating the differences between those averages and specific users and items.

Recommendations are then determined as follows: the estimate or prediction of a rating by user *u* of item *i*
= the global average rating + the user bias + the item bias.


### Movie Ratings

Below, the GBE is applied to sample movie ratings to make movie recommendations. First, the excel survey data is imported.

```{r import, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)

ratings <- read_excel("MovieRatings.xlsx")
```

Next, the average ratings for each `Critic` row (`user_biases`) and movie column (`item_biases`) are calculated.

```{r user-means}
user_biases <- data.frame(
  Critic = ratings$Critic,
  user_avg = round(rowMeans(ratings[2:7], na.rm = TRUE), digits = 2)
)

user_global_avg <- round(mean(user_biases$user_avg), digits = 2)
```

The global average for users is 4.03, which is used to calculate the user bias.

```{r user-bias}
user_biases <- user_biases |>
  mutate(user_bias = user_avg - user_global_avg)

user_biases
```

The process is similar for calculating the item bias, except the ratings table is pivoted longer to tidy the data first and the calculated global average for items is 3.87.

```{r item-bias}
ratings_tidy <- ratings |>
  pivot_longer(!Critic, names_to = "movies", values_to = "rating")

item_biases <- ratings_tidy |>
  group_by(movies) |>
  summarize(item_avg = round(mean(rating, na.rm = TRUE), digits = 2))

item_global_avg <- round(mean(item_biases$item_avg), digits = 2)

item_biases <- item_biases |>
  mutate(item_bias = item_avg - item_global_avg)

item_biases
```

These biases are joined onto the tidy dataset for easier calculations.

```{r joins}
gbe_ratings <- ratings_tidy |>
  filter(is.na(rating)) |>
  left_join(user_biases, by = "Critic") |>
  left_join(item_biases, by = "movies") |>
  mutate(predicted_rating = user_global_avg + user_bias + item_bias)

gbe_ratings
```

### Conclusion

The `predicted_rating` column of the new `gbe_ratings` dataframe should give reasonable predictions for each user-item pair. For example, each `Critic` could be given a list of movies they have not rated, filtered by any prediction over 4.
